% From https://www.sharelatex.com/learn/Typing_exams_in_LaTeX
% 
% To produce pdf run:
%   $ pdflatex paper.tex 
%

\documentclass[10pt]{exam}

\usepackage{listings}
\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
 
% These do what you would expect:
\printanswers
%\noprintanswers

\begin{document}
 
\title{SQL Workshop Week 3}
\author{UC Davis DSI}
\date{May 18, 2018}

\maketitle

\section{Review}

    These questions are meant to be more open ended. Challenge yourself by
    answering each one in two different ways. This requires three types of
    knowledge: knowledge of the data / domain (finance), knowledge of the
    tool (SQL), and critical thinking. This is the next step beyond following
    instructions.

\begin{questions}

\question What are the most profitable companies?

\begin{solution}

    1. We can compute standard metrics that don't exist in the database such as return on assets (ROA).
\texttt{DISTINCT} addresses a data quality issue by removing the duplicated
    rows.

\begin{lstlisting}
SELECT DISTINCT company_name, net_income / asset as return_on_assets
FROM company_info
ORDER BY  return_on_assets DESC
;
\end{lstlisting}

2.
Return on equity directly measures profitability for investors.

\begin{lstlisting}
SELECT cn.sandp_company_name, fr.return_on_equity
FROM company_name as cn
JOIN financial_ratios as fr
ON cn.ticker = fr.ticker
ORDER BY fr.return_on_equity DESC
;
\end{lstlisting}

\end{solution}


\question What are the most valuable sectors?

\begin{solution}
    This is a question about sectors, so you should be thinking about
    aggregating figures up to the sector level, which means using \texttt{GROUP
    BY sector}. Use \texttt{ORDER BY} so that the most valuable show up on
    top.

1.    Showing the assets in each sector in trillions of dollars rather than single units lets us
    readily read and compare them.

\begin{lstlisting}
SELECT sector, SUM(asset) / 1e12 as trillions
FROM company_info
GROUP BY sector
ORDER BY trillions DESC
;
\end{lstlisting}

2. This computes the average earnings per share in each industry

\begin{lstlisting}
SELECT sector, AVG(earning_per_share) as avg_earning_per_share
FROM company_info
GROUP BY sector
ORDER BY  avg_earning_per_share DESC
;
\end{lstlisting}


\end{solution}

\end{questions}
\section{Joins}

From Wikipedia:

    \begin{quote}
        The Standard Industrial Classification (SIC) is a system for
        classifying industries by a four-digit code. Established in the
        United States in 1937, it is used by government agencies to
        classify industry areas.
    \end{quote}

\begin{questions}

\subsection*{Single Joins}

\question Select the ticker, company name, and SIC code with corresponding description for
    companies that have SIC codes.

\begin{solution}
\begin{lstlisting}
SELECT ticker, company_name, sic_code, Description
FROM company_info
INNER JOIN sic
    ON company_info.sic_code = sic.SIC
;
\end{lstlisting}
\end{solution}


\question Select the ticker, company name, and SIC code with corresponding
    description (possibly NULL) for
    all companies in the \texttt{company\_info} table. 

\begin{solution}
\begin{lstlisting}
SELECT ticker, company_name, sic_code, Description
FROM company_info
LEFT JOIN sic
    ON company_info.sic_code = sic.SIC
;
\end{lstlisting}
\end{solution}

\question Which 4 digit SIC codes have the most companies?

\begin{solution}
\begin{lstlisting}
SELECT sic_code, Description, count(*) as cnt
FROM company_info
LEFT JOIN sic
    ON company_info . sic_code = sic . SIC
GROUP BY sic_code
ORDER BY cnt DESC
;
\end{lstlisting}
\end{solution}

\question Modify the query above to only return SIC codes with at least
5 companies.

\begin{solution}
\begin{lstlisting}
SELECT sic_code, Description, count(*) as cnt
FROM company_info
LEFT JOIN sic
    ON company_info . sic_code = sic . SIC
GROUP BY sic_code
HAVING cnt >= 5
ORDER BY cnt DESC
;
\end{lstlisting}
\end{solution}



\subsection*{Multiple Joins}

\question Use inner joins to produce a table with columns for company name,
state, SIC code, and SIC description.

\begin{solution}
\begin{lstlisting}
SELECT n.sandp_company_name, l.state,  i.sic_code, s.Description
FROM company_info as i
INNER JOIN company_name as n
    ON n.ticker = i.ticker
INNER JOIN sic as s
    ON i.sic_code = s.SIC
INNER JOIN company_locations as l
    ON l.ticker = i.ticker
;
\end{lstlisting}

Or you may prefer this syntax:

\begin{lstlisting}
SELECT n.sandp_company_name, l.state,  i.sic_code, s.Description
FROM company_info as i
    , company_name as n
    , sic as s 
    , company_locations as l 
WHERE n.ticker = i.ticker
AND i.sic_code = s.SIC
AND l.ticker = i.ticker
;
\end{lstlisting}
\end{solution}


\question Modify the above query to 
produce a table with columns for state, SIC
code, SIC description, and count of companies located in that state
with that SIC code.

\begin{solution}

    We no longer need company name, so we can remove the join to that
    table. Then we add a \texttt{GROUP BY} clause.

\begin{lstlisting}
SELECT l.state,  i.sic_code, s.Description, count(*) as cnt
FROM company_info as i
INNER JOIN sic as s
    ON i.sic_code = s.SIC
INNER JOIN company_locations as l
    ON l.ticker = i.ticker
GROUP BY i.sic_code, l.state
ORDER BY l.state ASC, cnt DESC
;
\end{lstlisting}
\end{solution}


\question Modify the query above to only return rows with a count
of at least 2 companies.

\begin{solution}
\begin{lstlisting}
SELECT l.state,  i.sic_code, s.Description, count(*) as cnt
FROM company_info as i
INNER JOIN sic as s
    ON i.sic_code = s.SIC
INNER JOIN company_locations as l
    ON l.ticker = i.ticker
GROUP BY i.sic_code, l.state
HAVING cnt >= 2
ORDER BY l.state ASC, cnt DESC
;
\end{lstlisting}
\end{solution}


\end{questions}
\end{document}
